Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-04-22 open_loop_planning.tex
[MM] robust adherence constraint

	\end{tikzpicture}
\caption{Schematic representation of the continuity and initialization equations associated with the $H+1$ propagated instances of the covariance matrix introduced in problem~\eqref{eq:DOCPopenloop}. The dashed rectangle highlights the $k$-th discretization step, at which the robust constraints are enforced. Each node corresponds to a specific covariance matrix, where the subscript denotes the current grid index and the superscript indicates the number of propagation steps since initialization. At each step, two key instances are emphasized: the matrix to be initialized at that step (red node), and the matrix that has been propagated over $H$ steps (green node), which is used in the evaluation of the constraint backoff.}
	%\caption{Schematic representation of the continuity and initialization Equations for the $H+1$ versions of the covariance matrix introduced in the discretized OCP. The dashed rectangle indicate the $k$-th step, in which the constraints are formulated. Each node represents a covariance matrix, labeled such that the subscript denotes the current grid step, while the superscript indicates the number of propagation steps it has undergone. At each step, two nodes are highlighted: the covariance matrix to be initialized (red node) and the matrix that has been propagated for $H$ steps (green node), which is used to formulate the robust constraints.}
	\label{fig:DOCPgrid}
\end{figure}

>>>>

	\end{tikzpicture}
	\caption{Schematic representation of the continuity and initialization equations associated with the $H+1$ propagated instances of the covariance matrix introduced in problem~\eqref{eq:DOCPopenloop}. The dashed rectangle highlights the $k$-th discretization step, at which the robust constraints are enforced. Each node corresponds to a specific covariance matrix, where the subscript denotes the current grid index and the superscript indicates the number of propagation steps since initialization. At each step, two key instances are emphasized: the matrix to be initialized at that step (red node), and the matrix that has been propagated over $H$ steps (green node), which is used in the evaluation of the constraint backoff.}
	%\caption{Schematic representation of the continuity and initialization Equations for the $H+1$ versions of the covariance matrix introduced in the discretized OCP. The dashed rectangle indicate the $k$-th step, in which the constraints are formulated. Each node represents a covariance matrix, labeled such that the subscript denotes the current grid step, while the superscript indicates the number of propagation steps it has undergone. At each step, two nodes are highlighted: the covariance matrix to be initialized (red node) and the matrix that has been propagated for $H$ steps (green node), which is used to formulate the robust constraints.}
	\label{fig:DOCPgrid}
\end{figure}
}{}

\ifdefstring{\version}{marco}{
\begin{figure}
	\centering
	\definecolor{mygreen}{RGB}{0,150,0}
	\begin{tikzpicture}[
		smallnode/.style={circle, draw, minimum size=3mm, inner sep=0pt},
		r_smallnode/.style={circle, draw=red, minimum size=3mm, inner sep=0pt},
		g_smallnode/.style={circle, draw=mygreen, minimum size=3mm, inner sep=0pt},
		r_hatnode/.style={
			circle, draw=red, pattern=north east lines, pattern color=red,
			minimum size=3mm, inner sep=0pt},
		g_hatnode/.style={
			circle, draw=mygreen, pattern=north east lines, pattern color=mygreen,
			minimum size=3mm, inner sep=0pt},
		every path/.style={->, thick},
		x=1.4cm, y=1.2cm
		]

		% ROW j = 0 (y=3)
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k-3}^0\)}] (P00) at (1,3) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k-2}^1\)}] (P11) at (2,3) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k-1}^2\)}] (P22) at (3,3) {};
		\node[g_hatnode, label={[label distance=-2pt]above:\(\bP_k^3\)}] (P33) at (4,3) {};
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k+1}^0\)}] (P44) at (5,3) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+2}^1\)}] (P55) at (6,3) {};

		% ROW j = 1 (y=2)
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k-2}^0\)}] (P10) at (2,2) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k-1}^1\)}] (P21) at (3,2) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_k^2\)}] (P32) at (4,2) {};
		\node[g_hatnode, label={[label distance=-2pt]above:\(\bP_{k+1}^3\)}] (P43) at (5,2) {};
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k+2}^0\)}] (P54) at (6,2) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+3}^1\)}] (P65) at (7,2) {};

		% ROW j = 2 (y=1)
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k-1}^0\)}] (P20) at (3,1) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_k^1\)}] (P31) at (4,1) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+1}^2\)}] (P42) at (5,1) {};
		\node[g_hatnode, label={[label distance=-2pt]above:\(\bP_{k+2}^3\)}] (P53) at (6,1) {};
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k+3}^0\)}] (P64) at (7,1) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+4}^1\)}] (P75) at (8,1) {};

		% ROW j = 3 (y=0)
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_k^0\)}] (P30) at (4,0) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+1}^1\)}] (P41) at (5,0) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+2}^2\)}] (P52) at (6,0) {};
		\node[g_hatnode, label={[label distance=-2pt]above:\(\bP_{k+3}^3\)}] (P63) at (7,0) {};
		\node[r_hatnode, label={[label distance=-2pt]above:\(\bP_{k+4}^0\)}] (P74) at (8,0) {};
		\node[smallnode, label={[label distance=-2pt]above:\(\bP_{k+5}^1\)}] (P85) at (9,0) {};

		% ARROWS
		\draw (P00) -- (P11);
		\draw (P11) -- (P22);
		\draw (P22) -- (P33);
		\draw (P22) -- (P33);

		\draw (P44) -- (P55);
		\draw[dashed,->] (P55) -- ++(1,0);

		\draw (P10) -- (P21);
		\draw (P21) -- (P32);
		\draw (P32) -- (P43);
		\draw (P54) -- (P65);
		\draw[dashed,->] (P65) -- ++(1,0);

		\draw (P20) -- (P31);
		\draw (P31) -- (P42);
		\draw (P42) -- (P53);
		\draw (P64) -- (P75);
		\draw[dashed,->] (P75) -- ++(1,0);

		\draw (P30) -- (P41);
		\draw (P41) -- (P52);
		\draw (P52) -- (P63);

		\draw (P74) -- (P85);
		\draw[dashed,->] (P85) -- ++(1,0);

		% Dashed box around column k (x=4)
		\draw[dashed, thick, rounded corners] (3.6,-0.5) rectangle (4.4,3.7);

		% Top axis labels
		\node at (0, 4.3) {\scriptsize \(\cdots\)};
		\node at (1, 4.3) {\scriptsize \(k-3\)};
		\node at (2, 4.3) {\scriptsize \(k-2\)};
		\node at (3, 4.3) {\scriptsize \(k-1\)};
		\node (knode) at (4, 4.3) {\scriptsize \(k\)};
		\node at (5, 4.3) {\scriptsize \(k+1\)};
		\node at (6, 4.3) {\scriptsize \(k+2\)};
		\node at (7, 4.3) {\scriptsize \(k+3\)};
		\node at (8, 4.3) {\scriptsize \(k+4\)};
		\node at (9, 4.3) {\scriptsize \(\cdots\)};

		% Wiggly arrow from label to top edge of dashed box
		\draw[->, thick, decorate, decoration={snake, amplitude=0.25mm, segment length=2.5mm}]
		(knode.south) -- (4, 3.8);

		% Legend box (top right)
		%\begin{scope}[shift={(6.5,2.5)}]
		%    \node[r_smallnode] (r) at (0,0) {};
		%    \node[anchor=west] at (0.4,0) {\scriptsize Initialized matrix \(\bP_k^0\)};
		%
		%    \node[g_smallnode] (g) at (0,-0.8) {};
		%    \node[anchor=west] at (0.4,-0.8) {\scriptsize Most propagated matrix \(\bP_k^3\)};
		%\end{scope}

	\end{tikzpicture}

	\caption{Cartoon representation (depicted for $H=3$) of the continuity and initialization equations associated with the $H+1$ propagated instances of the covariance matrix introduced in problem~\eqref{eq:DOCPopenloop}. The dashed rectangle highlights the $k$-th discretization step, at which the robust constraints are enforced. Each node corresponds to a specific covariance matrix, where the subscript denotes the current grid index and the superscript indicates the number of propagation steps since initialization. At each step, two key instances are emphasized: the matrix to be initialized at that step (red node), and the matrix that has been propagated over $H$ steps (green node), which is used in the evaluation of the constraint backoff.}
	%\caption{Schematic representation of the continuity and initialization Equations for the $H+1$ versions of the covariance matrix introduced in the discretized OCP. The dashed rectangle indicate the $k$-th step, in which the constraints are formulated. Each node represents a covariance matrix, labeled such that the subscript denotes the current grid step, while the superscript indicates the number of propagation steps it has undergone. At each step, two nodes are highlighted: the covariance matrix to be initialized (red node) and the matrix that has been propagated for $H$ steps (green node), which is used to formulate the robust constraints.}
	\label{fig:DOCPgrid}
\end{figure}
}{}

\subsection{Robust adherence limit constraint formulation}
The model employs a tire formulation that neglects combined slip effects, i.e., it does not account for the simultaneous utilization of longitudinal and lateral tyre forces. As a result, the optimal control problem must include an additional constraint to ensure that the total ground reaction forces remain within the bounds of the tyre's adherence ellipse.

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h(\bx,\bu) = \left[ \left( \frac{X(\bx,\bu)}{\mu_x} \right)^2+ \left( \frac{Y(\bx,\bu)}{\mu_y}\right)^2\right]\frac{1}{Z^2(\bx,\bu)}-1\leq0
\end{equation}
Here, $X$ and $Y$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z$ represents the total vertical load acting on the axle.

The gradient of the constraint w.r.t. the state vector $\bx$, denoted by $\na_{\bx}h(\bmu)$, is used to evaluate the standard deviation $\sigma$ associated with the constraint. This gradient can be computed using the chain rule, as follows:
\begin{equation}
	\na_{\bx}h(\bmu,\bu) = 
	\frac{\pd h}{\pd X}\at_{\bmu,\bu}\na_{\bx}X(\bmu,\bu) + 
	\frac{\pd h}{\pd Y}\at_{\bmu,\bu}\na_{\bx}Y(\bmu,\bu) + 
	\frac{\pd h}{\pd Z}\at_{\bmu,\bu}\na_{\bx}Z(\bmu,\bu)
\end{equation}
where the first element is zero because $X$ is a function of the input vector $\bu$ only.

For example, introducing a back-off term $\be=0.1$ in the constraint implies that the axle's adherence capacity can be exploited up to 90\% of its limit. This allows the optimizer to determine the most appropriate trade-off between the longitudinal force $X$ and the lateral force $Y$ --- often by reducing both components to some extent --- based on the specific requirements of the manoeuvre.
